\documentclass[11pt,openright]{book}

% Reality Under Inquiry - a self-contained, editable LaTeX book project
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{xspace}
\usepackage{geometry}
\geometry{paperwidth=6in,paperheight=9in,top=0.78in,bottom=0.75in,inner=0.86in,outer=0.70in,headsep=0.22in,footskip=0.40in}
\usepackage{xcolor}
\definecolor{covernavy}{HTML}{102A43}
\definecolor{ink}{HTML}{17212B}
\definecolor{muted}{HTML}{52606D}
\definecolor{teal}{HTML}{1F7A8C}
\definecolor{pale}{HTML}{F2F6F8}
\definecolor{warm}{HTML}{FBF7F1}
\definecolor{linegray}{HTML}{D9E2EC}
\color{ink}
\usepackage{setspace}
\setstretch{1.08}
\usepackage{parskip}
\setlength{\parskip}{0.55em plus 0.15em minus 0.08em}
\setlength{\emergencystretch}{2em}
\usepackage{amsmath,amssymb,mathtools}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{enumitem}
\setlist{topsep=0.35em,itemsep=0.15em,parsep=0.1em}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,calc,shapes.geometric}
\usepackage[most]{tcolorbox}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{csquotes}
\usepackage{xurl}
\usepackage{natbib}
\setcitestyle{authoryear,round,aysep={,},yysep={;}}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=teal,citecolor=teal,urlcolor=teal,plainpages=false,hypertexnames=false,pdftitle={Reality Under Inquiry},pdfauthor={Scott Brodie Forsyth}}
\usepackage{imakeidx}
\makeindex[title=Detailed Index,columns=2,intoc]

\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{1}
\makeatletter
\renewcommand*{\l@section}{\@dottedtocline{1}{1.5em}{3.5em}}
\makeatother

\newcommand{\RCR}{\textsc{RCR}\xspace}
\newcommand{\LCR}{\textsc{LCR}\xspace}
\newcommand{\doop}{\operatorname{do}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\claim}[1]{\textbf{#1}\index{#1}}

\titleformat{\chapter}[display]
  {\normalfont\bfseries\color{covernavy}}
  {\filleft\fontsize{11}{13}\selectfont\chaptertitlename\ \thechapter}
  {0.5em}
  {\filright\fontsize{25}{29}\selectfont}
  [\vspace{0.7ex}\titlerule]
\titlespacing*{\chapter}{0pt}{-5pt}{25pt}
\titleformat{\section}
  {\normalfont\Large\bfseries\color{covernavy}}{\thesection}{0.65em}{\raggedright}
\titleformat{\subsection}
  {\normalfont\large\bfseries\color{teal}}{\thesubsection}{0.65em}{\raggedright}
\titlespacing*{\section}{0pt}{2.2ex plus .4ex minus .2ex}{0.9ex}
\titlespacing*{\subsection}{0pt}{1.6ex plus .3ex minus .2ex}{0.6ex}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE]{\small\itshape\leftmark}
\fancyhead[RO]{\small\itshape\rightmark}
\fancyfoot[C]{\small\thepage}
\renewcommand{\headrulewidth}{0.35pt}
\renewcommand{\footrulewidth}{0pt}
\fancypagestyle{plain}{\fancyhf{}\fancyfoot[C]{\small\thepage}\renewcommand{\headrulewidth}{0pt}}
\renewcommand{\chaptermark}[1]{\markboth{\thechapter\quad #1}{}}
\renewcommand{\sectionmark}[1]{\markright{\thesection\quad #1}}

\newtcolorbox{principlebox}{
  enhanced,breakable,boxrule=0.35pt,arc=1.5pt,
  colback=pale,colframe=linegray,left=8pt,right=8pt,top=7pt,bottom=7pt,
  before skip=8pt,after skip=10pt
}
\newtcolorbox{objectionbox}{
  enhanced,breakable,boxrule=0.35pt,arc=1.5pt,
  colback=warm,colframe=linegray,left=8pt,right=8pt,top=7pt,bottom=7pt,
  before skip=8pt,after skip=10pt
}
\newtcolorbox{casebox}{
  enhanced,breakable,boxrule=0pt,arc=1.5pt,
  colback=covernavy!5,left=8pt,right=8pt,top=7pt,bottom=7pt,
  before skip=8pt,after skip=10pt
}

\newenvironment{chapterguide}{\begin{principlebox}\small\textbf{Chapter guide.}\ }{\end{principlebox}}
\newenvironment{discussion}{\begin{objectionbox}\small\textbf{Objections and discussion.}\ }{\end{objectionbox}}
\newenvironment{furtherreading}{\begin{casebox}\small\textbf{Further reading.}\ }{\end{casebox}}

\newcolumntype{Y}{>{\raggedright\arraybackslash}X}
\newcommand{\smalltable}{\small\renewcommand{\arraystretch}{1.15}}

\begin{document}

% Cover: intentionally title-only and typographically spaced.
\begin{titlepage}
  \pagecolor{covernavy}
  \color{white}
  \thispagestyle{empty}
  \vspace*{0.22\textheight}
  {\raggedright\fontsize{31}{35}\selectfont\bfseries REALITY\\UNDER INQUIRY\par}
  \vspace{1.6em}
  {\raggedright\fontsize{13}{18}\selectfont Toward an Epistemology and Ontology of Science\par}
  \vfill
  {\raggedright\large Scott Brodie Forsyth\par}
\end{titlepage}
\nopagecolor
\color{ink}

\frontmatter
\thispagestyle{empty}
\begin{center}
  {\Large\bfseries REALITY UNDER INQUIRY\par}
  \vspace{0.4em}
  {\large Toward an Epistemology and Ontology of Science\par}
  \vspace{1.2em}
  {\large Scott Brodie Forsyth\par}
\end{center}
\vfill
\noindent\rule{\textwidth}{0.35pt}\par
\small
This is an original philosophical work. Historical and scientific claims are
distinguished from interpretation, argument, and speculation. References are
given without invented quotations or page numbers; edition-specific pagination
should be checked against the copy used by a reader.
\vspace{1em}

Copyright \textcopyright\ 2026 Scott Brodie Forsyth. All rights reserved.
\clearpage

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
Science aims at knowledge of a reality that does not depend on our wishes,
languages, or institutions. Yet scientific knowledge is produced by historically
situated people through instruments, models, measurements, statistical methods,
and social systems that can both expose and conceal the world. This book asks
whether objectivity is possible under those conditions, and if so what it means.

The answer developed here is \emph{Reflexive Constraint Realism} (\RCR). Its
ontological thesis is that a mind-independent world contains entities,
processes, and relations that constrain what can happen. Its epistemological
thesis is that claims become objective to the extent that they survive
independent routes of constraint: reliable measurement, controlled variation,
intervention, cross-context transport, model comparison, critical disagreement,
and organized attempts to find error. The account is realist without being
naive, social without being relativist, and fallibilist without treating every
belief as equally good.

\RCR distinguishes four layers of scientific success. Correspondence concerns
whether a claim is about what exists or occurs. Structural adequacy concerns
whether a representation preserves relevant relations. Interventional adequacy
concerns whether manipulating a variable changes outcomes as the model says.
Pragmatic adequacy concerns whether a representation supports reliable inquiry
and action. These layers can come apart. A model may be useful while idealized;
an entity may be real while a theory about its deeper nature is wrong; and a
prediction may succeed for the wrong causal reason. The book therefore proposes
graded ontological commitment, explicit scope conditions, and a social norm of
public corrigibility.

The framework is reconstructed through logical empiricism, falsificationism,
Kuhn, Lakatos, Feyerabend, Bayesian and error-statistical approaches, inference
to the best explanation, realism and anti-realism, pragmatism, social and
feminist epistemology, naturalized epistemology, causal inference, model-based
science, and debates about explanation, laws, reduction, emergence, values,
replication, uncertainty, and demarcation. It is then tested against cases in
physics, biology, medicine, psychology, economics, climate science, and the
social sciences. The final chapters state the theory's axioms, inferential rules,
limits, objections, and an empirical research programme.

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}
The philosophy of science often appears as a sequence of positions: empiricism,
falsificationism, paradigms, research programmes, Bayesianism, realism,
anti-realism, and so on. That sequence is useful, but it can give the false
impression that the central task is to choose a camp. The more demanding task is
to understand why each camp saw something important, where it overreached, and
what a coherent account of objectivity would have to preserve.

This book is written in that spirit. It does not treat science as a single
algorithm. Nor does it treat the diversity of scientific practices as evidence
that there are no standards beyond local agreement. A laboratory experiment, a
climate model, a historical dataset, a randomized medical trial, and a causal
graph do not establish claims in the same way. They can nevertheless be judged
by shared questions: What is being claimed? What would count against it? Which
parts of the result depend on instruments, models, or social assumptions? Can
the claim travel to a new setting? What would an independent critic need in
order to find an error?

The framework proposed here is original in its organization and argumentative
role, not in the sense that every component has no precedent. The literature
already contains powerful ideas about intervention, structure, mechanisms,
social criticism, error, and realism. The contribution is to place these ideas
inside one account of objective inquiry, to specify their relations, and to
make their limits explicit.

The reader will find some formal notation. It is used as a tool for precision,
not as a display of technical status. Every symbol is introduced before it is
used. Where a historical claim is compressed, the bibliography points to a
primary source or a respected scholarly treatment. Where the book advances an
interpretation, it says so. Where it speculates, it marks the speculation.

\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}
Science is at once our most powerful way of learning about the world and a
human practice conducted through imperfect concepts, instruments, models,
institutions, and values. The central problem of this book is how those two
facts belong together. If science is objective, objectivity cannot mean
infallibility or a view from nowhere. If science is situated and revisable,
revisability cannot mean that evidence is powerless or that every description
is equally good.

The argument proceeds in three movements. The first reconstructs the principal
traditions in philosophy of science, from empiricism and logical positivism
through falsificationism, historical theories of change, Bayesian and
explanatory approaches, realism and anti-realism, pragmatism, and social and
feminist epistemologies. The second develops an account of observation,
models, measurement, causality, explanation, mechanisms, laws, emergence,
uncertainty, replication, and values. The third states and tests
\emph{Reflexive Constraint Realism} against cases in physics, biology,
medicine, psychology, climate science, economics, and the social sciences.

The result is a layered epistemology and ontology. It treats scientific claims
as scope-bound achievements whose warrant depends on evidence, invariance,
intervention, model transparency, and organized criticism. It also treats
their ontological force as graded: a successful pattern does not automatically
establish a hidden entity, but convergent structural and causal success can
justify increasingly strong commitments. The conclusion is a form of realism
disciplined by reflexivity, explicit uncertainty, and the possibility of
correction.

\tableofcontents

\mainmatter
\input{chapters/01-problem}
\input{chapters/02-empiricism}
\input{chapters/03-change}
\input{chapters/04-realism}
\input{chapters/05-observation}
\input{chapters/06-models}
\input{chapters/07-causality}
\input{chapters/08-social}
\input{chapters/09-reliability}
\input{chapters/10-ontology}
\input{chapters/11-epistemology}
\input{chapters/12-progress}
\input{chapters/13-cases}
\input{chapters/14-objections}

\backmatter
\setcounter{secnumdepth}{0}
\input{chapters/15-conclusion}
\input{chapters/16-glossary}
\input{chapters/17-audit}
\input{chapters/bibliography}

\printindex

\end{document}
