\chapter*{Bibliography}
\addcontentsline{toc}{chapter}{Bibliography}
\small
\begin{thebibliography}{99}

\bibitem[Anderson(2004)]{anderson2004}
Anderson, Elizabeth. 2004. \enquote{Uses of Value Judgments in Science: A General Argument, with Lessons from a Case Study of Feminist Research on Divorce.} \emph{Hypatia} 19(1): 1--24. https://doi.org/10.1111/j.1527-2001.2004.tb01266.x.

\bibitem[Armstrong(1983)]{armstrong1983}
Armstrong, D. M. 1983. \emph{What Is a Law of Nature?} Cambridge: Cambridge University Press.

\bibitem[Batterman(2002)]{batterman2002}
Batterman, Robert W. 2002. \emph{The Devil in the Details: Asymptotic Reasoning in Explanation, Reduction, and Emergence}. Oxford: Oxford University Press.

\bibitem[Bayes(1763)]{bayes1763}
Bayes, Thomas. 1763. \enquote{An Essay towards Solving a Problem in the Doctrine of Chances.} \emph{Philosophical Transactions of the Royal Society of London} 53: 370--418. https://doi.org/10.1098/rstl.1763.0053.

\bibitem[Bechtel and Abrahamsen(2005)]{bechtel2005}
Bechtel, William, and Adele Abrahamsen. 2005. \enquote{Explanation: A Mechanist Alternative.} \emph{Studies in History and Philosophy of Biological and Biomedical Sciences} 36(2): 421--441. https://doi.org/10.1016/j.shpsc.2005.03.010.

\bibitem[Bertrand and Mullainathan(2004)]{bertrandmullainathan2004}
Bertrand, Marianne, and Sendhil Mullainathan. 2004. \enquote{Are Emily and Greg More Employable than Lakisha and Jamal? A Field Experiment on Labor Market Discrimination.} \emph{American Economic Review} 94(4): 991--1013. https://doi.org/10.1257/0002828042002561.

\bibitem[Bogen and Woodward(1988)]{bogenwoodward1988}
Bogen, James, and James Woodward. 1988. \enquote{Saving the Phenomena.} \emph{The Philosophical Review} 97(3): 303--352. https://doi.org/10.2307/2185445.

\bibitem[Cartwright(1983)]{cartwright1983}
Cartwright, Nancy. 1983. \emph{How the Laws of Physics Lie}. Oxford: Clarendon Press.

\bibitem[Cartwright(1999)]{cartwright1999}
Cartwright, Nancy. 1999. \emph{The Dappled World: A Study of the Boundaries of Science}. Cambridge: Cambridge University Press.

\bibitem[Chakravartty(2007)]{chakravartty2007}
Chakravartty, Anjan. 2007. \emph{A Metaphysics for Scientific Realism: Knowing the Unobservable}. Cambridge: Cambridge University Press.

\bibitem[Craver(2007)]{craver2007}
Craver, Carl F. 2007. \emph{Explaining the Brain: Mechanisms and the Mosaic Unity of Neuroscience}. Oxford: Clarendon Press.

\bibitem[Douglas(2009)]{douglas2009}
Douglas, Heather. 2009. \emph{Science, Policy, and the Value-Free Ideal}. Pittsburgh: University of Pittsburgh Press.

\bibitem[Duhem(1906)]{duhem1906}
Duhem, Pierre. 1906. \emph{La theorie physique: son objet et sa structure}. Paris: Chevalier et Riviere. English translation, 1954, \emph{The Aim and Structure of Physical Theory}, translated by Philip P. Wiener. Princeton: Princeton University Press.

\bibitem[Einstein(1905)]{einstein1905}
Einstein, Albert. 1905. \enquote{Ueber die von der molekularkinetischen Theorie der Waerme geforderte Bewegung von in ruhenden Fluessigkeiten suspendierten Teilchen.} \emph{Annalen der Physik} 17: 549--560. English translation: \enquote{On the Movement of Small Particles Suspended in Stationary Liquids Required by the Molecular-Kinetic Theory of Heat.}

\bibitem[Feyerabend(1975)]{feyerabend1975}
Feyerabend, Paul. 1975. \emph{Against Method: Outline of an Anarchistic Theory of Knowledge}. London: New Left Books.

\bibitem[Fine(1984)]{fine1984}
Fine, Arthur. 1984. \enquote{The Natural Ontological Attitude.} In \emph{Scientific Realism}, edited by Jarrett Leplin, 83--107. Berkeley: University of California Press.

\bibitem[Franklin and Gosling(1953)]{franklingosling1953}
Franklin, Rosalind E., and Raymond G. Gosling. 1953. \enquote{Molecular Configuration in Sodium Thymonucleate.} \emph{Nature} 171: 740--741. https://doi.org/10.1038/171740a0.

\bibitem[Frigg and Hartmann(2006)]{frigghartmann2006}
Frigg, Roman, and Stephan Hartmann. 2006. \enquote{Models in Science.} In \emph{The Stanford Encyclopedia of Philosophy}, edited by Edward N. Zalta. Revised 2020. https://plato.stanford.edu/entries/models-science/.

\bibitem[Giere(1988)]{giere1988}
Giere, Ronald N. 1988. \emph{Explaining Science: A Cognitive Approach}. Chicago: University of Chicago Press.

\bibitem[Giere(2004)]{giere2004}
Giere, Ronald N. 2004. \enquote{How Models Are Used to Represent Reality.} \emph{Philosophy of Science} 71(5): 742--752. https://doi.org/10.1086/425063.

\bibitem[Goldman(1999)]{goldman1999}
Goldman, Alvin I. 1999. \emph{Knowledge in a Social World}. Oxford: Oxford University Press.

\bibitem[Hacking(1983)]{hacking1983}
Hacking, Ian. 1983. \emph{Representing and Intervening: Introductory Topics in the Philosophy of Natural Science}. Cambridge: Cambridge University Press.

\bibitem[Haraway(1988)]{haraway1988}
Haraway, Donna. 1988. \enquote{Situated Knowledges: The Science Question in Feminism and the Privilege of Partial Perspective.} \emph{Feminist Studies} 14(3): 575--599. https://doi.org/10.2307/3178066.

\bibitem[Harding(1991)]{harding1991}
Harding, Sandra. 1991. \emph{Whose Science? Whose Knowledge? Thinking from Women's Lives}. Ithaca: Cornell University Press.

\bibitem[Harman(1965)]{harman1965}
Harman, Gilbert. 1965. \enquote{The Inference to the Best Explanation.} \emph{The Philosophical Review} 74(1): 88--95. https://doi.org/10.2307/2183532.

\bibitem[Hempel and Oppenheim(1948)]{hempeloppenheim1948}
Hempel, Carl G., and Paul Oppenheim. 1948. \enquote{Studies in the Logic of Explanation.} \emph{Philosophy of Science} 15(2): 135--175. https://doi.org/10.1086/286983.

\bibitem[Hempel(1965)]{hempel1965}
Hempel, Carl G. 1965. \emph{Aspects of Scientific Explanation and Other Essays in the Philosophy of Science}. New York: Free Press.

\bibitem[Hernan and Robins(2020)]{hernanrobins2020}
Hernan, Miguel A., and James M. Robins. 2020. \emph{Causal Inference: What If}. Boca Raton: Chapman \& Hall/CRC.

\bibitem[Holland(1986)]{holland1986}
Holland, Paul W. 1986. \enquote{Statistics and Causal Inference.} \emph{Journal of the American Statistical Association} 81(396): 945--960. https://doi.org/10.1080/01621459.1986.10478354.

\bibitem[Ioannidis(2005)]{ioannidis2005}
Ioannidis, John P. A. 2005. \enquote{Why Most Published Research Findings Are False.} \emph{PLoS Medicine} 2(8): e124. https://doi.org/10.1371/journal.pmed.0020124.

\bibitem[Kitcher(1993)]{kitcher1993}
Kitcher, Philip. 1993. \emph{The Advancement of Science: Science without Legend, Objectivity without Illusions}. New York: Oxford University Press.

\bibitem[Kitcher(2001)]{kitcher2001}
Kitcher, Philip. 2001. \emph{Science, Truth, and Democracy}. New York: Oxford University Press.

\bibitem[Kuhn(1962)]{kuhn1962}
Kuhn, Thomas S. 1962. \emph{The Structure of Scientific Revolutions}. Chicago: University of Chicago Press. 2nd ed., with postscript, 1970.

\bibitem[Lakatos(1970)]{lakatos1970}
Lakatos, Imre. 1970. \enquote{Falsification and the Methodology of Scientific Research Programmes.} In \emph{Criticism and the Growth of Knowledge}, edited by Imre Lakatos and Alan Musgrave, 91--196. Cambridge: Cambridge University Press.

\bibitem[Lakatos(1978)]{lakatos1978}
Lakatos, Imre. 1978. \emph{The Methodology of Scientific Research Programmes: Philosophical Papers, Volume 1}, edited by John Worrall and Gregory Currie. Cambridge: Cambridge University Press.

\bibitem[Laudan(1977)]{laudan1977}
Laudan, Larry. 1977. \emph{Progress and Its Problems: Toward a Theory of Scientific Growth}. Berkeley: University of California Press.

\bibitem[Laudan(1983)]{laudan1983}
Laudan, Larry. 1983. \enquote{The Demise of the Demarcation Problem.} In \emph{Physics, Philosophy and Psychoanalysis}, edited by Robert S. Cohen and Larry Laudan, 111--127. Dordrecht: D. Reidel.

\bibitem[Longino(1990)]{longino1990}
Longino, Helen E. 1990. \emph{Science as Social Knowledge: Values and Objectivity in Scientific Inquiry}. Princeton: Princeton University Press.

\bibitem[Lucas(1976)]{lucas1976}
Lucas, Robert E., Jr. 1976. \enquote{Econometric Policy Evaluation: A Critique.} In \emph{The Phillips Curve and Labor Markets}, edited by Karl Brunner and Allan H. Meltzer, 19--46. Carnegie-Rochester Conference Series on Public Policy, vol. 1. Amsterdam: North-Holland.

\bibitem[Mayo(1996)]{mayo1996}
Mayo, Deborah G. 1996. \emph{Error and the Growth of Experimental Knowledge}. Chicago: University of Chicago Press.

\bibitem[Medical Research Council(1948)]{mrc1948}
Medical Research Council. 1948. \enquote{Streptomycin Treatment of Pulmonary Tuberculosis: A Medical Research Council Investigation.} \emph{British Medical Journal} 2(4582): 769--782. https://doi.org/10.1136/bmj.2.4582.769.

\bibitem[Merton(1942)]{merton1942}
Merton, Robert K. 1942. \enquote{The Normative Structure of Science.} In \emph{The Sociology of Science: Theoretical and Empirical Investigations}, 1973, 267--278. Chicago: University of Chicago Press.

\bibitem[Michelson and Morley(1887)]{michelsonmorley1887}
Michelson, Albert A., and Edward W. Morley. 1887. \enquote{On the Relative Motion of the Earth and the Luminiferous Ether.} \emph{American Journal of Science} 34(203): 333--345.

\bibitem[Morgan and Morrison(1999)]{morganmorrison1999}
Morgan, Mary S., and Margaret Morrison, eds. 1999. \emph{Models as Mediators: Perspectives on Natural and Social Science}. Cambridge: Cambridge University Press.

\bibitem[Nagel(1961)]{nagel1961}
Nagel, Ernest. 1961. \emph{The Structure of Science: Problems in the Logic of Scientific Explanation}. New York: Harcourt, Brace \& World.

\bibitem[Open Science Collaboration(2015)]{osc2015}
Open Science Collaboration. 2015. \enquote{Estimating the Reproducibility of Psychological Science.} \emph{Science} 349(6251): aac4716. https://doi.org/10.1126/science.aac4716.

\bibitem[Oppenheim and Putnam(1958)]{oppenheimputnam1958}
Oppenheim, Paul, and Hilary Putnam. 1958. \enquote{Unity of Science as a Working Hypothesis.} In \emph{Minnesota Studies in the Philosophy of Science}, vol. 2, edited by Herbert Feigl, Michael Scriven, and Grover Maxwell, 3--36. Minneapolis: University of Minnesota Press.

\bibitem[Peirce(1877--1878)]{peirce1878}
Peirce, Charles S. 1877--1878. \enquote{The Fixation of Belief} and \enquote{How to Make Our Ideas Clear.} In \emph{The Essential Peirce}, vol. 1, edited by Nathan Houser and Christian Kloesel, 109--123, 124--141. Bloomington: Indiana University Press, 1992.

\bibitem[Pearl(2009)]{pearl2009}
Pearl, Judea. 2009. \emph{Causality: Models, Reasoning, and Inference}. 2nd ed. Cambridge: Cambridge University Press.

\bibitem[Perrin(1913)]{perrin1913}
Perrin, Jean. 1913. \emph{Les Atomes}. Paris: Alcan.

\bibitem[Popper(1959)]{popper1959}
Popper, Karl. 1959. \emph{The Logic of Scientific Discovery}. London: Hutchinson. English translation of the 1934 German edition.

\bibitem[Quine(1951)]{quine1951}
Quine, W. V. O. 1951. \enquote{Two Dogmas of Empiricism.} \emph{The Philosophical Review} 60(1): 20--43. https://doi.org/10.2307/2266637.

\bibitem[Quine(1969)]{quine1969}
Quine, W. V. O. 1969. \enquote{Epistemology Naturalized.} In \emph{Ontological Relativity and Other Essays}, 69--90. New York: Columbia University Press.

\bibitem[Reichenbach(1938)]{reichenbach1938}
Reichenbach, Hans. 1938. \emph{Experience and Prediction: An Analysis of the Foundations and the Structure of Knowledge}. Chicago: University of Chicago Press.

\bibitem[Reiss and Sprenger(2014)]{reisssprenger2014}
Reiss, Julian, and Jan Sprenger. 2014. \enquote{Scientific Objectivity.} In \emph{The Stanford Encyclopedia of Philosophy}, edited by Edward N. Zalta. Revised 2020. https://plato.stanford.edu/entries/scientific-objectivity/.

\bibitem[Rubin(1974)]{rubin1974}
Rubin, Donald B. 1974. \enquote{Estimating Causal Effects of Treatments in Randomized and Nonrandomized Studies.} \emph{Journal of Educational Psychology} 66(5): 688--701. https://doi.org/10.1037/h0037350.

\bibitem[Salmon(1984)]{salmon1984}
Salmon, Wesley C. 1984. \emph{Scientific Explanation and the Causal Structure of the World}. Princeton: Princeton University Press.

\bibitem[Sellars(1963)]{sellars1963}
Sellars, Wilfrid. 1963. \emph{Science, Perception and Reality}. London: Routledge \& Kegan Paul.

\bibitem[Stanford(2006)]{stanford2006}
Stanford, Kyle P. 2006. \emph{Exceeding Our Grasp: Science, History, and the Problem of Unconceived Alternatives}. New York: Oxford University Press.

\bibitem[Thomson(1897)]{thomson1897}
Thomson, J. J. 1897. \enquote{Cathode Rays.} \emph{The London, Edinburgh, and Dublin Philosophical Magazine and Journal of Science} 44(269): 293--316. https://doi.org/10.1080/14786449708621070.

\bibitem[van Fraassen(1980)]{vanfraassen1980}
van Fraassen, Bas C. 1980. \emph{The Scientific Image}. Oxford: Clarendon Press.

\bibitem[Watson and Crick(1953)]{watsoncrick1953}
Watson, James D., and Francis H. C. Crick. 1953. \enquote{A Structure for Deoxyribose Nucleic Acid.} \emph{Nature} 171: 737--738. https://doi.org/10.1038/171737a0.

\bibitem[Weber(1949)]{weber1949}
Weber, Max. 1949. \enquote{Objectivity in Social Science and Social Policy.} In \emph{The Methodology of the Social Sciences}, translated and edited by Edward A. Shils and Henry A. Finch, 49--112. New York: Free Press.

\bibitem[Woodward(2003)]{woodward2003}
Woodward, James. 2003. \emph{Making Things Happen: A Theory of Causal Explanation}. New York: Oxford University Press.

\bibitem[Worrall(1989)]{worrall1989}
Worrall, John. 1989. \enquote{Structural Realism: The Best of Both Worlds?} \emph{Dialectica} 43(1--2): 99--124. https://doi.org/10.1111/j.1746-8361.1989.tb00933.x.

\bibitem[IPCC(2023)]{ipcc2023}
Intergovernmental Panel on Climate Change. 2023. \emph{Climate Change 2023: Synthesis Report}. Geneva: IPCC. https://www.ipcc.ch/report/ar6/syr/.

\end{thebibliography}

